Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-12-31 astronum_2018/astronum-xrb.tex
more referee addressing

Flame spreading was first modeled in a series of calculations
employing an algorithm from atmospheric science, where vertical
hydrostatic equilibrium is enforced and wide-aspect ratio zones give a
horizontal CFL number that is large enough to obtain long timescale
evolution~\cite{cavecchi:2012}.
These calculations display key features of the propagation
mechanism, namely a balance between hydrodynamics and the pure
flame physics. The flame proceeds mainly via conduction, thus being a
deflagration. The flame front, however, is not vertical but inclined,
the angle with the horizontal being very small: $\theta \sim
H/R_{\rm{Ro}} \sim 10^{-3}$. Here $H$ is the scale height and
$R_{\rm{Ro}}$ is the Rossby radius. It is the hydrodynamical balance
between the Coriolis force, gravity and pressure gradients that
determines the Rossby radius and therefore the inclination angle of the
flame front. The extended surface of the flame front is what leads to
propagation speeds of order $10^5 $~km/s, in good agreement with
observations. Due to its dependence on the Coriolis force, the flame
speed changes across the surface of the NS (being faster at the
equator), leading to potentially detectable effects if the rise of the
burst is observed with enough resolution
\cite{art-2015-cavecchi-etal}.

@@

Figure~\ref{fig:xrb} shows the flame structure for an initial 2D run
at 0.007~s.  The burning layer is only about 10~m high.  Underneath is
\isot{Ni}{56}, representative of the underlying neutron star, and
above it is a very low density ($10^{-4}~\gcc$) buffer.  Only a
portion of the vertical extent is shown.  The top panel shows the
temperature structure.  The initial perturbation reached to $2\times
10^4~\mathrm{cm}$, so at this point, the extent of the burnt region
has nearly doubled.  The flame structure is quite wide, since there
are multiple burning stages beyond just helium to carbon, and this
shows up in the composition.  This is seen more clearly in the next
panel which shows the mean molecular weight of the nuclei ($\bar{A}$).
We see that the further behind the head of the flame we are, the heavier
the ash nuclei.  We also see the head of the flame
appears lifted off of the base of the burning layer---unburned
material underlies the spreading flame.  This is
reminiscent of the flame structure seen in the earlier simulations of
\cite{cavecchi:2012}.  The third panel shows the energy generation rate.
It is strongest just above the neutron star surface, behind the flame.
We again see that the head of the flame is detached from the base
of the atmosphere, with the strongest burning at the head of the flame
slightly higher up in the atmosphere.   Finally, the bottom
panel shows the velocity through the plane of the simulation---this is
induced by the Coriolis force as the burning fluid expands and begins to spread laterally.
A tight hurricane structure has setup as a result of the flame
spreading.  This is the geostrophic balance discussed in
\cite{SPIT_ETAL02}.    These calculations are ongoing and will be
the subject of a detailed study in the near future.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the mean molecular
  weight of the ash (second), the nuclear energy generation rate
  (third), and the velocity out of the simulation plane (bottom).  The
  out-of-plane velocity contours indicate where the Coriolis force has
  created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.  }
\end{figure}

>>>>

Flame spreading was first modeled in a series of calculations
employing an algorithm from atmospheric science, where vertical
hydrostatic equilibrium is enforced and wide-aspect ratio zones give a
horizontal CFL number that is large enough to obtain long timescale
evolution~\cite{cavecchi:2012}.  These calculations display key
features of the propagation mechanism, namely a balance between
hydrodynamics and the pure flame physics. The flame proceeds mainly
via conduction, thus being a deflagration. The flame front, however,
is not vertical but inclined, the angle with the horizontal being very
small: $\theta \sim H/R_{\rm{Ro}} \sim 10^{-3}$. Here $H$ is the scale
height and $R_{\rm{Ro}}$ is the Rossby radius. It is the
hydrodynamical balance between the Coriolis force, gravity and
pressure gradients that determines the Rossby radius and therefore the
inclination angle of the flame front. The extended surface of the
flame front is what leads to propagation speeds of order $10^5 $~km/s,
in good agreement with observations. Due to its dependence on the
Coriolis force, the flame speed changes across the surface of the NS
(being faster at the equator), leading to potentially detectable
effects if the rise of the burst is observed with enough resolution
\cite{art-2015-cavecchi-etal}.  We note that no calculations have
considered the effects of the flame interacting with preexisting
turbulence ahead of the flame from simmering that may be occurring
there, as we characterized in \cite{xrb3}.  Flame-turbulence
interactions have long been seen as a means to accelerate a flame in
Type Ia supernovae, so future studies should look at turbulence in the
layer.  Capturing turbulence, of course, requires high resolution, so
the ideas described here will be needed.

@@

Figure~\ref{fig:xrb} shows the flame structure for an initial 2D run
at 0.007~s.  Vertical profiles at a few radial distances from the
origin are shown for the same quantities in
Figure~\ref{fig:xrb-profiles}.  The burning layer is only about 10~m
high.  Underneath is \isot{Ni}{56}, representative of the underlying
neutron star, and above it is a very low density ($10^{-4}~\gcc$)
buffer.  Only a portion of the vertical extent is shown.  The top
panel shows the temperature structure.  The initial perturbation
reached to $2\times 10^4~\mathrm{cm}$, so at this point, the extent of
the burnt region has nearly doubled.  The flame structure is quite
wide, since there are multiple burning stages beyond just helium to
carbon, and this shows up in the composition.  This is seen more
clearly in the next panel which shows the mean molecular weight of the
nuclei ($\bar{A}$).  We see that the further behind the head of the
flame we are, the heavier the ash nuclei.  We also see the head of the
flame appears lifted off of the base of the burning layer---unburned
material underlies the spreading flame.  This is reminiscent of the
flame structure seen in the earlier simulations of
\cite{cavecchi:2012}.  The third panel shows the energy generation
rate.  It is strongest just above the neutron star surface, behind the
flame.  We again see that the head of the flame is detached from the
base of the atmosphere, with the strongest burning at the head of the
flame slightly higher up in the atmosphere.  Finally, the bottom panel
shows the velocity through the plane of the simulation---this is
induced by the Coriolis force as the burning fluid expands and begins
to spread laterally.  A tight hurricane structure has setup as a
result of the flame spreading.  This is the geostrophic balance
discussed in \cite{SPIT_ETAL02}.  These calculations are ongoing and
will be the subject of a detailed study in the near future.

\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the mean molecular
  weight of the ash (second), the nuclear energy generation rate
  (third), and the velocity out of the simulation plane (bottom).  The
  out-of-plane velocity contours indicate where the Coriolis force has
  created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.  }
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{xrb_profiles}
\caption{\label{fig:xrb_profiles} Vertical profiles at several laterial distances
  from the vertical axis through the plots shown in Figure~\ref{fig:xrb}.  These
  profiles sample the region behind and ahead of the flame.}
\end{figure}
